Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2024-04-27 report/sections/methods.Rmd
rearrange markdown sections

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{colortbl}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage[utf8]{inputenc}
\usepackage{icomma}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{csquotes}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
  pdflang={en-EN},
  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev\textsuperscript{3}}
\date{}

\shorttitle{Modelling writing hesitations}

\authornote{

Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}

\abstract{%
Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning. Our results cannot be explained by the serial but are in line with the cascading view of writing.
}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).

Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech. Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language. Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012). This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g. Clark \& Fox Tree, 2002). Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).

The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models. On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006). For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019). The same is generally observed for spoken utterances (E.-K. Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001). We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted. Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.

Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991). The intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured. Processing may be serial. This means that the processing cascade operates on one unit at a time. Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002). This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).

This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002). However this account is not consistent with what we know about writing. Consider these two observations:

First, utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process. This is well known for spoken production (e.g. Bock \& Ferreira, 2014). For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al. (2019){]}. For example, in Roeser et al. (2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset. This is not what one would expect if writers plan sentences (or even phrases) in full before production onset. In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al. (2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.

Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid. For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration. Also Rønneberg et al. (2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control. Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991). Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016). Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two data-generating processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).

For example, Roeser et al. (2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling. In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach. Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions. As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes. For similar approaches see Almond et al. (2012), Baaijen et al. (2012), Chenu et al. (2014), Guo et al. (2018), Hall et al. (2022), Li (2021), and Van Waes et al. (2021).

The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. Rønneberg et al. (2022) coined this possibility the \enquote{process-disruption hypothesis}.

In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al. (2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).

Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production. Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).

Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023). Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016. These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).

Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021). In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks. Instead text composition provides data from both inter and intra-sentence transitions. Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts. Full modelling details can be found in Appendix \textbackslash ref\{statistical-models; we summarise the most important model properties in the following. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data. In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.

>>>>

In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts. Full modelling details can be found in Appendix \\ref{statistical-models; we summarise the most important model properties in the following. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data. In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.

The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling. Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes. These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs. This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.

We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts. We discussed advantages with this approach in the introduction section. These datasets include samples from populations with various writing experience and languages (e.g. young / L2 writers, students) performing different writing tasks (e.g. essays, syntheses). Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.

## Data sets

TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2. For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say "unpublished". I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk). Was the GUNNEXP2 sample Norwegian? Should we include information like time limit or text length in the table?

Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; descriptions below. Dataset used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GUNNEXP2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, PLanTra data are published in @rossetti2022text and described in @rossetti2022s. In C2L1 Norwegian 6th graders composed argumentative essays. In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text. In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced. The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each. The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability. The GUNNEXP2 data .... [MARK, please add a sentence here to describe the production context].

For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203]. LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].

\blandscape

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
                Source = c("Rønneberg et al. (2022)",
                           "Torrance et al. (2016)",
                           "Torrance et al. (n.d.)",
                           "Rossetti and Van Waes (2022b)",
                           "Vandermeulen, Steendam, et al. (2020)", 
                           "Torrance and Ofstad (n.d)"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog",
                           "EyeWrite"),
                `Writing task` = c("Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis", 
                                 "?????"),
                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, "????"),
                 Sample = c("6th graders", "Secondary school students", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "???? Norway"),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch",
                               "???? Norwegian")) %>% 
  arrange(Dataset) %>% 
  apa_table(align = c("l", "p{3cm}", "p{2cm}", "p{2.5cm}", "r", "r", "p{2.5cm}", "p{2.25cm}", "p{2.5cm}"), 
            escape = FALSE, 
            font_size = "footnotesize",
            caption = "Datasets in brief.")

#                Manipulation = c("--", 
#                               "Weak decoders / control; texted masked / unmasked",
#                               "Text produced in L1 / L2",
#                               "Pre / post test with training on plain language principles", 
#                              "Various topics and genres",
#                               "masked / unmasked")
```

\elandscape

## Data extraction

From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful [@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}. In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions. At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis. Also we removed transitions that were followed by an editing operation.

```{r keyloc, results='asis'}

table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
       Description = c("Transitions between any letter",
                       "Keypress after space followed by any letter",
                       "Keypress following a space preceding any letter"),
       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d. T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.", 
                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))

apa_table(table, caption = "Transition location classification.",
            align = c("p{3cm}", "p{5cm}", "p{6cm}"), 
            escape = FALSE, 
            font_size = "footnotesize",
            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace.")

```

```{r reductionfunctions}
source("../scripts/get_data_summary.R")

# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../../data/spl2.csv", n_samples = n_samples) 
gunnexp2 <- get_gunnexp2("../../data/gunnexp2.csv", n_samples = n_samples)
```

We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data). We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant). We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or too long ($\ge$ 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}. From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available). This was done to reduce the computation time of the Bayesian models require to complete sampling. For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets. Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us. The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.

```{r datareduction, results='asis'}
c2l1$ds <- "c2l1"
cato$ds <- "cato"
lift$ds <- "lift"
plantra$ds <- "plantra"
spl2$ds <- "spl2"
gunnexp2$ds <- "gunnexp2"

extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
          mutate(ds = .x$ds)) %>% 
  mutate(across(where(is.numeric), ~signif(.*100, 2))) %>% 
  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")")),
         across(c(`<50ms`, `>30,000ms`), ~str_replace(., "0 \\(0\\)", "--"))) %>% 
  rename(`$\\le$ 50 msecs` = `<50ms`,
         `$\\ge$ 30 secs` = `>30,000ms`)

random_sample <- 
  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
          mutate(ds = .x$ds)) %>% 
    mutate(across(where(is.numeric), ~signif(.*100, 2))) %>%
    unite("keep", starts_with("keep"), sep = " (") %>% 
    mutate(across(keep, ~str_c(., ")")),
          across(keep, ~str_replace(., "100 \\(0\\)", "--"))) %>%
  pivot_wider(names_from = location, values_from = keep) %>% 
  relocate(`within word`)

datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
  relocate(ds) %>% 
  mutate(across(ds, ~recode_factor(., 
                "c2l1" = "C2L1",
                "cato" = "CATO",
                "gunnexp2" = "GUNNEXP2",
                "lift" = "LIFT",
                "plantra" = "PLanTra",
#                "spl2_shift" = "SPL2 (shift + C)", 
                "spl2" = "SPL2",
                .ordered = T
                ))) %>% 
  arrange(ds) %>% 
  rename(`Dataset` = ds)

apa_table(datareduction, 
          escape = FALSE,  
          placement = "bp!",
          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used). Standard error in parentheses.",
          align =c("l", rep("r", 5)),
          col_spanners = list(" " = 1,
                              "Extreme values in \\%" = c(2, 3),
                              "Randomly sampled data in \\%" = c(4, 6)))

```

## Bayesian statitical modelling
